Early Warning Signals and the Prosecutor's Fallacy

### 2012-05-16 fallacy.tex
fallacy updated

a model with a linear death rate and density dependent birth rate that drives
the Allee effect at low densities and limits growth at high densities.  The
key feature of this model is the alternate stable states introduced by this effect;
other functional forms for Eq.~\ref{master} could serve equally well for these
simulations~\citep[see \emph{e.g.}][]{Scheffer2001}. 
Though this system can be forced through a saddle-node bifurcation by
increasing the death rate, in these simulations all parameters
are held constant and no bifurcation occurs.  Consequently we do not 
anticipate an early warning signal of an approaching bifurcation.
The simulation starts from the positive equilibrium population size.
Though the chance of a transition across the Allee threshold in any 
given time step are small, given enough time this system will eventually
experience such a rare event driving the population extinct.  We evolved
each replicate 50,000 time steps, sampling the system every 50 time steps.  
In this time window 266 of the 1,000 replicates experience population collapse.  
Nine of these replicates can be seen in Figure~\ref{fig:replicate_crashes},
as an illustration.

  \begin{figure}
    \begin{center}
      \includegraphics[width=6in]{replicate_crashes.pdf}
    \end{center}
    \caption{Example trajectories for nine of the 266 replicates that 
             experience a critical transition by chance during the course
             of the simulation.}
    \label{fig:replicate_crashes}
  \end{figure}

@@

To test for the presence of a warning signal in these indicators we computed
values of Kendall's $\tau$ for both indicators for each of the 266 replicates.
Kendall's $tau$ takes values in $(-1, 1)$, and is frequently used to identify
an increasing trend ($\tau > 0 )$ in early warning signals.~\citep{Dakos2008, Dakos2011}.
indicate an increasing trend in the warning signal.  The distribution of $\tau$ 
values observed across these replicates is shown in Figure~\ref{fig:indicator}.
More positive values indicate stronger trends. The common statistical significance
test used for Kendall's $tau$ assumes independence of the variables to achieve
a normal distribution, this isn't satisfied for indicators coming from a time 
series and computed with a sliding window.  For a more robust comparison, we 
compare the distribution of this statistic from all the simulations to the 
distribution conditioned on experiencing a chance transition to the alternative
stable state, that is, from the replicates which are known to crash.  To avoid 
an effect of sample size the timeseries are all chosen to be the same length.  
As these examples have been selected from a system that is not approaching a 
bifurcation, an indicator that does not suffer from the bias of Prosecutor's fallacy
should have the same distribution in both cases.   Annotated, parallelized code to 
replicate the analysis and produce the figures is provided in the appendix.

@@

\begin{equation}
  \frac{dX}{dt} = r - x^2
\end{equation}
Where we introduce an explicit parameter $\theta$ to account for rescaling
to the units of the data,

@@

We have shown that selecting datasets conditional on having experienced sudden 
noise-induced transitions can lead to patterns that are more likely to resemble
early warning signals typically associated with critical slowing down before 
a bifurcation event.  Purely noise-induced transitions are distinctly separate 
mechanism from critical slowing-down phenomena, in which the transition occurs
as the result of a single or series of extremely unlikely events rather than
a slow but regular degradation of the system stability.

It seems tempting to argue that the bias towards positive detection in historical
examples is in fact valuable -- each of these systems did indeed collapse, so the fact that they show increased 
probability to exhibit warning signals could be taken as a successful detection.
Unfortunately, this reasoning is flawed. At the moment the forecast is made,
these systems are no more likely to crash than their counterparts from the sample
that did not collapse, just as a fair coin is still a fair coin after a string of
consecutive ``heads.''  Taking a closer look at the patterns involved we can
begin to understand why common indicators such as autocorrelation and variance
can be misleading.

@@

Because we have selected our data by conditioning
on the fact that they have experienced such a crash, these excursions appear
more frequently than we would expect by chance in the observations of a given
stable system over some finite time window.  That biased selection is the
heart of the Prosecutor's Fallacy.  In these cases, the presence of increasing
autocorrelation or increasing variance does not represent the critical slowing
down process.

@@

%% Comments on reproducibility?

>>>>

a model with a linear death rate and density dependent birth rate that drives
the Allee effect at low densities and limits growth at high densities.  The
key feature of this model is the alternate stable states introduced by this effect;
other functional forms for Eq.~\ref{master} could serve equally well for these
simulations~\citep[see \emph{e.g.}][]{Scheffer2001}. 
Though this system can be forced through a saddle-node bifurcation by
increasing the death rate, in these simulations all parameters
are held constant and no bifurcation occurs.  Consequently we do not 
anticipate an early warning signal of an approaching bifurcation.
The simulation starts from the positive equilibrium population size.
Though the chance of a transition across the Allee threshold in any 
given time step are small, given enough time this system will eventually
experience such a rare event driving the population extinct.  We evolved
each replicate 50,000 time steps, sampling the system every 50 time steps.  
In this time window 266 of the 1,000 replicates experience population collapse.  
To keep the examples of comperable sample size, we focus on a section of the data
500 time points prior to the system approaching the transition.  
Nine of these replicates can be seen in Figure~\ref{fig:replicate_crashes},
as an illustration.

  \begin{figure}
    \begin{center}
      \includegraphics[width=6in]{replicate_crashes.pdf}
    \end{center}
    \caption{Example trajectories for nine of the 266 replicates that 
             experience a critical transition by chance during the course
             of the simulation, showing the window of 500 points prior to
             crossing the threshold which is used in the analysis.  A wobble
             at the unstable critical point is visible at the end of several 
             trajectories.}
    \label{fig:replicate_crashes}
  \end{figure}

@@

To test for the presence of a warning signal in these indicators we computed
values of Kendall's $\tau$ for both indicators for each of the 266 replicates.
Kendall's $tau$ takes values in $(-1, 1)$, and is frequently used to identify
an increasing trend ($\tau > 0 )$ in early warning signals.~\citep{Dakos2008, Dakos2011}.
indicate an increasing trend in the warning signal.  The distribution of $\tau$ 
values observed across these replicates is shown in Figure~\ref{fig:indicator}.
More positive values indicate stronger trends. The common statistical significance
test used for Kendall's $tau$ assumes independence of the variables to achieve
a normal distribution, this isn't satisfied for indicators coming from a time 
series and computed with a sliding window.  For a more robust comparison, we 
compare the distribution of this statistic from all the simulations to the 
distribution conditioned on experiencing a chance transition to the alternative
stable state, that is, from the replicates which are known to crash.  To avoid 
an effect of sample size the timeseries are all chosen to be the same length.  
As these examples have been selected from a system that is not approaching a 
bifurcation, an indicator that does not suffer from the bias of Prosecutor's fallacy
should have the same distribution in both cases.   Annotated, parallelized code to 
replicate the analysis and produce the figures is provided in the appendix. 
(Currently at \href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).

@@

\begin{equation}
  \frac{dX}{dt} = r - x^2
\end{equation}
where we introduce an explicit parameter $\theta$ to account for rescaling
to the units of the data,

@@

Historical examples of systems that have undergone sudden transitions seem
to offer an invaluable opportunity to test methods for the detection of
early warning signals that could have forseen the collapse.  However,
such data must be used with care, as the process of selecting systems
already known to have collapsed opens up the possibility that these
transitions were the result of rare chance events rather than examples
of the more deterministic tipping point behaviour we wish to detect.
When we think about forecasting potential collapse in a single system,
these purely-chance transitions may be rare enough to ignore.  When we
comb the historical record for any available data on sudden transitions,
our odds of finding these chance transitions may be much higher.

We have shown that systems which undergo such rare transitions look
statistically different from their counterparts that do not, even though
they are driven by the same stochastic process.  In particular, such
conditionally selected examples are more likely to show signs associated
with an early warning of an approaching tipping point, such as increasing
variance or increasing autocorrelation, as measured by Kendall's $\tau$.
This increases the risk of false positives -- cases in which a warning
signal being tested appears to have successfully detected an oncoming
tipping point, when in fact the example is falsely accused, coming
instead from a stable system. Figure 2 shows the extent of this bias,
where many of the chance crashes show values of $\tau$ that are 
significantly larger than those observed in the otherwise identical
replicates that did not experience a chance transition.

\subsection{Chance transitions are false positives for early warning signals}

It seems tempting to argue that the bias towards positive detection
in historical examples is valuable -- each of these systems did
indeed collapse, so the fact that they show increased probability
to exhibit warning signals could be taken as a successful detection.
Unfortunately this is not the case. At the moment the forecast is made,
these systems are no more likely to crash than their counterparts from
the sample that did not collapse, just as a fair coin is still a fair
coin after a string of consecutive ``heads.''  Taking a closer look at
the patterns involved we can begin to understand why common indicators
such as autocorrelation and variance can be misleading.

@@

\subsection{Comparing to the model-based method}

In our experiment, the model-based estimate of early warning signals
appears more robust than the summary statistics, producing the same
estimates on both the conditionally selected replicates as on a random 
sample of the replicates.  This is a consequence of the more rigid 
specifications that come with a model-based approach -- the pattern
expected is less general than any increase in variance or autocorrelation,
but instead must be one that matches its approximation of the saddle-node
bifurcation. This observation highlights the difference between the pattern
driving the false positive trends in increasing variance and increasing
autocorrelation and the pattern anticipated in the saddle-node model. This
should not however be taken as evidence that the model-based approach is 
immune to the bias of the Prosecutor's Fallacy.

@@
